---
title: "Structural Equation Modeling with R using lavaan"
author: "Alexander M. Schoemann"
date: "SPSP 2015"
output:
  beamer_presentation:
    slide_level: 2
    includes:
      in_header: header.tex
---

## What is SEM?

- AKA:
    - Simultaneous Equations
    - Covariance Structure Analysis
    - Path Analysis
    - Confirmatory Factor Analysis (CFA) 

## What is a model?

- A model is a set of relations between observed and latent variables that can be represented by a set of equations.
- Or...
- A useful representation of reality

> Box: "essentially, all models are wrong, but some are useful"

## Latent variables

- A latent variable:
    - is a construct that is hypothesized to exist but is not directly measurable
    - causes behaviors that are directly measurable; these behaviors are called indicators of the latent variable
    - is defined as everything that is in common between its manifest (observed) indicators 


## SEM in R

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(dev = 'pdf')
```

- There are (at least) four packages that fit SEM models in R
    - lavaan, OpenMx, sem, and lava


## lavaan
> a free open-source, but commercial-quality package for latent variable modeling.

    - From lavaan.org
    
- Features
    - Full support for mean structures and multiple groups
    - Several estimators available (including ML, GLS, WLS...)
    - Standard and robust standard errors and test statistics (bootstrapping too!)
    - Missing data handling through FIML
    - Linear and non-linear inequality constraints
    - Support for categorical data (and mixture of binary, ordered and continuous observed variables)

## lavaan syntax

- lavaan specifies relationships based on a path diagram
    - Every arrow in the path diagram is a line of syntax in lavaan

- lavaan syntax uses a few different operators to specify relationships between variables
    - =~ is used for factor loadings
        - ```Positive =~ Great```
    - ~~ is used for variance and covariances
        - ```Great ~~ Great```
    - ~ is used for regressions 
    - \* is used to fix a parameter to a specific value
        - ```Positive ~~ 1*Positive```

## lavaan: Example
- Two factor CFA based on the msq daa (from the psych package)

```{r, echo=FALSE, include=FALSE}
library(lavaan)
library(psych)
library(semTools)
library(semPlot)
```

```{r}
mod <- '
energetic =~ active + vigorous + wakeful + lively 
tense =~ intense + jittery + fearful
'

fit <- cfa(mod, data = msq)
```

## lavaan: Example

By default lavaan uses a marker variable method of scale setting

```{r, echo = TRUE}
semPaths(fit, "est", nCharNodes = 0)
```


## lavaan: Example
```{r}
fit1 <- cfa(mod, data = msq, std.lv = TRUE)
```
The std.lv option will use a fixed factor method of identification

## lavaan: Example
```{r, echo = FALSE}
semPaths(fit1, "est", nCharNodes = 0)
```

## lavaan: Example
- Model results can be inspected with summary, nested models can be compared with anova

```
Estimator                                         ML
  Minimum Function Test Statistic              323.554
  Degrees of freedom                                13
  P-value (Chi-square)                           0.000

User model versus baseline model:

  Comparative Fit Index (CFI)                    0.970
  Tucker-Lewis Index (TLI)                       0.952

Loglikelihood and Information Criteria:

  Loglikelihood user model (H0)             -26806.206
  Loglikelihood unrestricted model (H1)     -26644.430

  Number of free parameters                         22
  Akaike (AIC)                               53656.413
  Bayesian (BIC)                             53794.285
  Sample-size adjusted Bayesian (BIC)        53724.379
```

## lavaan: Example

```
Root Mean Square Error of Approximation:

  RMSEA                                          0.078
  90 Percent Confidence Interval          0.071  0.086
  P-value RMSEA <= 0.05                          0.000

Standardized Root Mean Square Residual:

  SRMR                                           0.048

```

## lavaan: Example
```
                   Estimate  Std.err  Z-value  P(>|z|)
Latent variables:
  energetic =~
    active            0.798    0.012   64.708    0.000
    vigorous          0.639    0.011   56.253    0.000
    wakeful           0.614    0.013   46.768    0.000
    lively            0.785    0.012   67.324    0.000
  tense =~
    intense           0.681    0.019   35.674    0.000
    jittery           0.413    0.016   25.531    0.000
    fearful           0.121    0.008   14.830    0.000

Covariances:
  energetic ~~
    tense             0.579    0.017   34.129    0.000

```

## lavaan: Example Multiple Groups 
- Two factor CFA based on the msq daa (from the psych package)
    - Compare the msq with the revised msq

```{r}
mod <- '
energetic =~ active + vigorous + wakeful + lively 
tense =~ intense + jittery + fearful
'

fitg <- cfa(mod, data = msq, group = "scale")
```

## lavaan: Example Multiple Groups 
- Parameters can constrained with group.equal option

```{r}
fitgW <- cfa(mod, data = msq, group = "scale",
            group.equal = "loadings")

anova(fitg, fitgW)
```

## lavaan: Example Multiple Groups 
- Or invariance testing can be conducted using the measurementInvariance function
    - In the semTools package

```{r, results = "hide"}
measurementInvariance(mod, data = msq, group = "scale")
```

## lavaan: Example Multiple Groups
```

Model 1: configural invariance:
    chisq        df    pvalue       cfi     rmsea       bic 
  343.506    26.000     0.000     0.970     0.080 53254.619 

Model 2: weak invariance (equal loadings):
    chisq        df    pvalue       cfi     rmsea       bic 
  352.357    31.000     0.000     0.969     0.073 53222.196 

[Model 1 versus model 2]
  delta.chisq      delta.df delta.p.value     delta.cfi 
        8.851         5.000         0.115         0.000 

Model 3: strong invariance (equal loadings + intercepts):
    chisq        df    pvalue       cfi     rmsea       bic 
  376.702    36.000     0.000     0.967     0.070 53205.266 

[Model 1 versus model 3]
  delta.chisq      delta.df delta.p.value     delta.cfi 
       33.195        10.000         0.000         0.002 

[Model 2 versus model 3]
  delta.chisq      delta.df delta.p.value     delta.cfi 
       24.344         5.000         0.000         0.002         
```

## lavaan: Example Categorical Indicators 
- Two factor CFA based on the ability data (from the psych package)

```{r, include = FALSE}
ability <- data.frame(ability)
```

```{r}
modCat <- '
reason =~ reason.4 + reason.16 + reason.17 + reason.19
letter =~ letter.7 + letter.33 + letter.34 + letter.58
matrix =~ matrix.45 + matrix.46 + matrix.47 + matrix.55
rotate =~ rotate.3 + rotate.4 + rotate.6 + rotate.8
'

fitCat <- cfa(modCat, data = ability, std.lv = TRUE, 
              ordered = names(ability))
```

## lavaan: Example Categorical Indicators 
```{r, echo = FALSE}
semPaths(fitCat, intercept = FALSE)
```

## lavaan: Extensions

- There are many packages providing additional functionality to lavaan:
    - semPlot - Path diagrams for lavaan (used to make the above plots)
    - semTools - useful functions for SEM in R
        - Includes functions for latent interactions, multiple imputation and more
    - simsem - simulation package for SEM 
    - lavaan.survey - apply survey weights, clustering corrections and other corrections for lavaan models
    - Onyx - graphical user interface for SEM (http://onyx.brandmaier.de/)
    
## lavaan: Resources

- The lavaan website is extremely helpful (lavaan.org)
    - It include a tutorial in lavaan
    - Links to examples from popular SEM books in lavaan
    - Link to a discussion board about lavaan


## Thank you!

- Questions?
- email: schoemanna@ecu.edu


- Slides from today at: 

