---
title: "Structural Equation Modeling with R using lavaan"
author: "Alexander M. Schoemann"
date: "SPSP 2015"
output:
  beamer_presentation:
    slide_level: 2
    includes:
      in_header: header.tex
---

## What is SEM?

- AKA:
    - Simultaneous Equations
    - Covariance Structure Analysis
    - Path Analysis
    - Confirmatory Factor Analysis (CFA) 

## SEM in R

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(dev = 'pdf')
```

- There are (at least) four packages that fit SEM models in R
    - lavaan, OpenMx, sem, and lava


## lavaan
> a free open-source, but commercial-quality package for latent variable modeling.

    - From lavaan.org
    
- lavaan is extensively tested and updated
    - Bugs are quickly reported and corrected

- To install lavaan:
```
install.packages("lavaan")
```

    
## lavaan    
- Features
    - Full support for mean structures and multiple groups
    - Several estimators available (including ML, GLS, WLS...)
    - Standard and robust standard errors and test statistics (bootstrapping too!)
    - Missing data handling through FIML
    - Linear and non-linear inequality constraints
    - Support for categorical data (and mixture of binary, ordered and continuous observed variables)

## lavaan syntax

- lavaan specifies relationships based on a path diagram
    - Every arrow in the path diagram is a line of syntax in lavaan

- lavaan syntax uses a few different operators to specify relationships between variables
    - =~ is used for factor loadings
        - ```energetic =~ active```
    - ~~ is used for variance and covariances
        - ```active ~~ active```
    - ~ is used for regressions 
    - \* is used to fix a parameter to a specific value
        - ```energetic ~~ 1*energetic```

## lavaan syntax

- Users can also specify models with code from other software and fit them with lavaan
    - mplus2lavaan (in the lavaan package)
    - lisrel2lavaan (in the semTools package)
    - Onyx (graphical user interface)

## lavaan: Example
- Two factor CFA based on the msq data (from the psych package)

```{r, echo=FALSE, include=FALSE}
library(lavaan)
library(psych)
library(semTools)
library(semPlot)
```

```{r}
mod <- '
energetic =~ active + vigorous + wakeful + lively 
negative =~  jittery + nervous + scared
'

fit <- cfa(mod, data = msq)
```

## lavaan: Example

By default lavaan uses a marker variable method of scale setting

```{r, echo = TRUE}
semPaths(fit, "est", nCharNodes = 0)
```


## lavaan: Example
```{r}
fit1 <- cfa(mod, data = msq, std.lv = TRUE)
```
- The std.lv option will use a fixed factor method of identification

## lavaan: Example
```{r, echo = FALSE}
semPaths(fit1, "est", nCharNodes = 0)
```

## lavaan: Example
- Model results can be inspected with summary, nested models can be compared with anova

```
  Estimator                                         ML
  Minimum Function Test Statistic              354.351
  Degrees of freedom                                13
  P-value (Chi-square)                           0.000

User model versus baseline model:

  Comparative Fit Index (CFI)                    0.969
  Tucker-Lewis Index (TLI)                       0.950

Loglikelihood and Information Criteria:

  Loglikelihood user model (H0)             -25561.582
  Loglikelihood unrestricted model (H1)     -25384.406

```

## lavaan: Example

```
  Number of free parameters                         15
  Akaike (AIC)                               51153.164
  Bayesian (BIC)                             51246.989
  Sample-size adjusted Bayesian (BIC)        51199.326

Root Mean Square Error of Approximation:

  RMSEA                                          0.083
  90 Percent Confidence Interval          0.075  0.090
  P-value RMSEA <= 0.05                          0.000

Standardized Root Mean Square Residual:

  SRMR                                           0.076
  
```

## lavaan: Example
```
                   Estimate  Std.err  Z-value  P(>|z|)
Latent variables:
  energetic =~
    active            0.796    0.012   63.720    0.000
    vigorous          0.632    0.011   55.166    0.000
    wakeful           0.615    0.013   46.630    0.000
    lively            0.792    0.012   67.459    0.000
  negative =~
    jittery           0.429    0.014   29.904    0.000
    nervous           0.571    0.013   42.769    0.000
    scared            0.282    0.009   32.220    0.000

Covariances:
  energetic ~~
    negative          0.181    0.018    9.853    0.000
```

## lavaan: Example Multiple Groups 
- Two factor CFA based on the msq data (from the psych package)
    - Compare the msq with the revised msq

```{r}
mod <- '
energetic =~ active + vigorous + wakeful + lively 
negative =~  jittery + nervous + scared
'

fitg <- cfa(mod, data = msq, group = "scale")
```

## lavaan: Example Multiple Groups 
- Parameters can constrained with group.equal option

```{r}
fitgW <- cfa(mod, data = msq, group = "scale",
            group.equal = "loadings")

anova(fitg, fitgW)
```

## lavaan: Example Multiple Groups 
- Or invariance testing can be conducted using the measurementInvariance function
    - In the semTools package

```{r, results = "hide"}
measurementInvariance(mod, data = msq, group = "scale")
```

## lavaan: Example Multiple Groups
```
Model 1: configural invariance:
    chisq        df    pvalue       cfi     rmsea       bic 
  380.753    26.000     0.000     0.968     0.084 51380.620 

Model 2: weak invariance (equal loadings):
    chisq        df    pvalue       cfi     rmsea       bic 
  396.410    31.000     0.000     0.967     0.078 51355.002 

[Model 1 versus model 2]
  delta.chisq      delta.df delta.p.value     delta.cfi 
       15.657         5.000         0.008         0.001 

Model 3: strong invariance (equal loadings + intercepts):
    chisq        df    pvalue       cfi     rmsea       bic 
  403.573    36.000     0.000     0.967     0.073 51320.889 

[Model 1 versus model 3]
  delta.chisq      delta.df delta.p.value     delta.cfi 
       22.820        10.000         0.011         0.001 

[Model 2 versus model 3]
  delta.chisq      delta.df delta.p.value     delta.cfi 
        7.163         5.000         0.209         0.000     
```

## lavaan: Example Categorical Indicators 
- Two factor CFA based on the ability data (from the psych package)

```{r, include = FALSE}
ability <- data.frame(ability)
```

```{r}
modCat <- '
reason =~ reason.4 + reason.16 + reason.17 + reason.19
letter =~ letter.7 + letter.33 + letter.34 + letter.58
matrix =~ matrix.45 + matrix.46 + matrix.47 + matrix.55
rotate =~ rotate.3 + rotate.4 + rotate.6 + rotate.8
'

fitCat <- cfa(modCat, data = ability, std.lv = TRUE, 
              ordered = names(ability))
```

## lavaan: Example Categorical Indicators 
```{r, echo = FALSE}
semPaths(fitCat, "est", intercept = FALSE, nCharNodes = 6)
```

## lavaan: Extensions

- There are many packages providing additional functionality to lavaan:
    - semPlot - Path diagrams for lavaan (used to make the above plots)
    - semTools - useful functions for SEM in R
        - Includes functions for latent interactions, multiple imputation and more
    - simsem - simulation package for SEM 
    - lavaan.survey - apply survey weights, clustering corrections and other corrections for lavaan models
    - Onyx - graphical user interface for SEM (http://onyx.brandmaier.de/)
    
## lavaan: Resources

- The lavaan website is extremely helpful (http://lavaan.org)
    - It include a tutorial in lavaan
    - Links to examples from popular SEM books in lavaan
    - Link to a discussion board about lavaan


## Thank you!

- Questions?
- email: schoemanna@ecu.edu


- Slides and code: https://sites.google.com/site/alexandermschoemann/supplementarymaterials

